---
title: "KIRC Multi-Omics Dashboard"
format:
  dashboard:
    theme: cosmo
    css: styles.css
    orientation: rows
    nav-buttons:
      - icon: github
        href: https://github.com/Lexiyao/kidney-cancer-multiomics
---

```{r}
#| label: setup
#| include: false
library(plotly)
library(ggplot2)
library(survival)
source("../R/constants.R")
source("../R/functions_dashboard.R")

# Frozen research targets. fn_dashboard_read() returns NULL when the
# `targets-store` release asset has not been restored, so every card below
# degrades to a stated gap rather than taking the site render down.
subtypes_df      <- fn_dashboard_read("subtypes_df")
km_src           <- fn_dashboard_read("km_subtype_df")
survival_metrics <- fn_dashboard_read("survival_metrics")
bap1_auroc       <- fn_dashboard_read("bap1_auroc")
concordance      <- fn_dashboard_read("concordance")
sanity_table     <- fn_dashboard_read("sanity_table")
rna_mat          <- fn_dashboard_read("rna_mat")
```

## Row {height=14%}

::: {.card title="Every number on this page is FROZEN"}
All figures here come from the **frozen `curatedTCGAData` snapshot 20160128
(hg19)**, integrated once and cached in the `_targets` store. They **never
update** — a weekly re-run reproduces them byte-for-byte. The project's only
live-updating numbers are on the [Live GDC](live-gdc.qmd) page, and a
frozen-snapshot count is **not** interchangeable with a current GDC portal
count: the two differ materially in both directions.
:::

## Row {height=22%}

```{r}
#| content: valuebox
#| title: "Cohort n — FROZEN snapshot (RNA ∩ Methyl ∩ CNV)"
fn_dashboard_valuebox(
  if (is.null(subtypes_df)) NULL else nrow(subtypes_df),
  icon = "people", colour = "primary", fmt = "%d"
)
```

```{r}
#| content: valuebox
#| title: "Survival C-index (held-out, Cox)"
fn_dashboard_valuebox(survival_metrics$cindex$cox,
                      icon = "activity", colour = "info")
```

```{r}
#| content: valuebox
#| title: "BAP1-from-expression AUROC (held-out)"
fn_dashboard_valuebox(bap1_auroc$heldout_auroc,
                      icon = "cpu", colour = "success")
```

```{r}
#| content: valuebox
#| title: "MOFA vs SNF concordance (ARI)"
fn_dashboard_valuebox(concordance$ari,
                      icon = "diagram-3", colour = "secondary", fmt = "%.2f")
```

## Row {height=32%}

### Column {width=40%}

```{r}
#| title: "MOFA subtype sizes (frozen)"
if (is.null(subtypes_df)) {
  htmltools::div(class = "pending-note",
                 fn_dashboard_pending("the MOFA subtype sizes"))
} else {
  tab <- as.data.frame(table(subtype = subtypes_df$subtype))
  plot_ly(tab, x = ~subtype, y = ~Freq, type = "bar",
          marker = list(color = DASHBOARD_ACCENT)) |>
    layout(xaxis = list(title = "MOFA subtype"),
           yaxis = list(title = "Samples"))
}
```

### Column {width=60%}

```{r}
#| title: "Overall survival by MOFA subtype — DESCRIPTIVE, not a model output"
if (is.null(km_src)) {
  htmltools::div(class = "pending-note",
                 fn_dashboard_pending("the Kaplan-Meier curves by subtype"))
} else {
  km <- survfit(Surv(time, status) ~ subtype, data = km_src)
  plot_ly(fn_km_curve_df(km), x = ~time, y = ~surv, color = ~strata,
          type = "scatter", mode = "lines", line = list(shape = "hv")) |>
    layout(xaxis = list(title = "Time (days)"),
           yaxis = list(title = "Overall survival", range = c(0, 1)))
}
```

## Row {height=32%}

### Column {.tabset width=60%}

```{r}
#| title: "Driver-gene expression by subtype"
# `rna_mat` is the top-5000-variable slice, so a driver gene can be absent from
# it. An empty selection used to reach facet_wrap() and abort the whole site
# render ("Faceting variables must have at least one value."), and a partial one
# used to be shown without saying what was missing. Both cases are decided by
# fn_driver_gene_frame() and covered by tests/testthat/test-dashboard.R, because
# neither is reachable in a render without the store.
if (is.null(rna_mat) || is.null(subtypes_df)) {
  htmltools::div(class = "pending-note",
                 fn_dashboard_pending("the driver-gene expression view"))
} else {
  gd <- fn_driver_gene_frame(rna_mat, subtypes_df)
  if (nrow(gd$frame) == 0L) {
    htmltools::div(class = "pending-note", fn_driver_gene_gap(gd))
  } else {
    p_gene <- ggplot(gd$frame, aes(x = subtype, y = expr, fill = subtype)) +
      geom_boxplot(outlier.size = 0.4) +
      facet_wrap(~gene, scales = "free_y") +
      labs(x = "MOFA subtype", y = "log2(x+1) normalised expression") +
      theme_minimal(base_size = 10) +
      theme(legend.position = "none")
    gap_note <- fn_driver_gene_note(gd$dropped)
    htmltools::tagList(
      ggplotly(p_gene),
      if (is.null(gap_note)) NULL else {
        htmltools::div(class = "dashboard-note", gap_note)
      }
    )
  }
}
```

```{r}
#| title: "Positive controls — including the one that FAILS"
if (is.null(sanity_table)) {
  htmltools::div(class = "pending-note",
                 fn_dashboard_pending("the positive-control results table"))
} else {
  # fn_pass_span() + escape = FALSE, exactly as index.qmd renders the SAME
  # table. Plain "FAIL (RED)" body text on this card is how the m1-m4 red light
  # stops looking like a red light (dashboard/styles.css states the rule).
  knitr::kable(
    data.frame(
      Check    = sanity_table$check,
      Verdict  = fn_pass_span(sanity_table$passed),
      Evidence = sanity_table$detail,
      check.names = FALSE, stringsAsFactors = FALSE
    ),
    escape = FALSE, row.names = FALSE
  )
}
```

```{r}
#| title: "What is NOT on this page"
#| echo: false
htmltools::div(
  class = "dashboard-note",
  htmltools::HTML(paste0(
    "<p>The value boxes above carry no confidence intervals and no ",
    "<em>p</em>-values, because at this cohort size an interval is the honest ",
    "half of every one of them. Read them with the qualifications on the ",
    "<a href='survival.qmd'>Survival</a> and ",
    "<a href='factors.qmd'>MOFA Factors</a> pages, not on their own.</p>",
    "<p>Nothing here is filled in when the frozen store is missing: a card ",
    "with no store prints a stated gap. Every value box and figure on this ",
    "page is a live read from the store; the numbers in the red-light card ",
    "beside them are prose quoted from the committed run transcripts under ",
    "<code>docs/results/</code>, and carry their run id.</p>"
  ))
)
```

### Column {width=40%}

::: {.card title="Read the red light before the value boxes"}
**The merged-methylation 4-cluster positive control FAILS, and it stays
failing here.** (`k = 4` is a design choice taken from the m1–m4 mRNA
*expression* subtypes of TCGA Nature 2013, which publishes no four
DNA-methylation strata — so this is a finding about the assay merge, not a
failed replication.) `methyl_mat` is `cbind(HM27, HM450)` on common CpGs with **no batch
correction** (cohort split HM27 214 / HM450 310), and the adjusted Rand index
against the assay split is **0.583** versus a 0.25 veto. Worse for the
optimistic reading: the strata are *not* recovered inside a single platform
either — within-arm 4-means silhouettes are HM27 0.0858 and HM450 0.0489,
**below** the merged 0.1197 — so the merge manufactures the apparent structure
rather than diluting real structure. No batch correction was applied and none
will be (only 3 cases overlap the platforms; the probe sets differ).

**What survives that.** The MOFA *subtypes* are platform-clean (ARI 0.0058
against the assay split), so the subtype views above are not pictures of the
confound. Mutation frequencies (VHL 44.8 %, PBRM1 30.5 %, SETD2 10.1 %,
BAP1 8.6 %) and ccA/ccB separation (rho −0.354) touch no methylation matrix at
all and PASS.

**The BAP1 survival control passes its direction test and is underpowered.**
HR 1.584 (95 % CI 0.967–2.595, *p* = 0.068). Roughly 470 events would be needed
for 80 % power; the mutation subset supplies **147, of which 18 are in the
mutant arm** — recorded by run 31375702141, not inferred from the interval.
That is neither a confirmation nor a negative finding, and it is not reported
as either.

**The held-out C-index is not fully out-of-sample.** MOFA was fitted on all 524
cases, including the rows later held out, so the latent axes were defined with
the test rows contributing. MOFA never sees an outcome — this is transductive
optimism, not label leakage — but the figure bounds the supervised component
only, and there is no external validation cohort.
:::
