---
title: "MOFA2 Factors"
format:
  html:
    page-layout: full
---

```{r}
#| label: setup
#| include: false
library(plotly)
library(ggplot2)
source("../R/constants.R")
source("../R/functions_dashboard.R")

# Frozen research targets. fn_dashboard_read() returns NULL when the release
# asset has not been restored, so this page degrades to stated gaps instead of
# taking the whole site render down (see R/functions_dashboard.R).
mofa_varexp     <- fn_dashboard_read("mofa_varexp")
mofa_factors    <- fn_dashboard_read("mofa_factors")
factor_platform <- fn_dashboard_read("factor_platform")
methyl_platform <- fn_dashboard_read("methyl_platform")
has_store       <- !is.null(mofa_varexp) && !is.null(factor_platform)
```

::: {.provenance .provenance-frozen}
[Frozen]{.provenance-label} Every number on this page comes from the **frozen
`curatedTCGAData` snapshot 20160128 (hg19)**, integrated once and cached in the
`_targets` store. It **never updates**. The only live-updating numbers in this
project are on the [Live GDC](live-gdc.qmd) page, and frozen-snapshot counts do
not match the current GDC portal in either direction.
:::

## Read this before reading any factor

MOFA2 was fitted on three views — log-transformed normalised RNA, merged
HM27∪HM450 methylation M-values, and GISTIC-thresholded CNV — over the
**524-case** main cohort. Mutation is *not* a view; it is an external label used
only to interpret factors.

The methylation view is `cbind(HM27, HM450)` on common CpGs with **no batch
correction**, and the cohort splits **HM27 214 / HM450 310**. Several factors
are therefore substantially the assay split rather than biology. Every factor
below is shown with its measured rank-biserial AUC against that split
(0.5 = no platform information, BH-adjusted across the 15 factors); a factor
with `q ≤ 0.05` is marked [CONFOUNDED by assay]{.verdict-red} and must not be
read as a biological axis.

**The same merge fails a literature positive control, and this is the page the
rest of the site sends you to for the confound — so it says so here rather than
only on the [Dashboard](dashboard.qmd).** The merged-methylation 4-cluster
anchor is [RED]{.verdict-red} on this snapshot: the merged 4-means partition
tracks the assay at ARI **0.583** against a 0.25 veto, and the within-platform
silhouettes (HM27 0.0858, HM450 0.0489) are *both below* the merged 0.1197, so
the merge manufactures the structure rather than diluting it. It is pinned as an
expected red — a *new* red fails the job, and this anchor starting to pass fails
it too. `k = 4` is a design choice taken from the m1–m4 mRNA **expression**
subtypes (TCGA Nature 2013), which reports no four DNA-methylation strata, so
this red is a finding about the assay merge and not a failed replication. See
the [Dashboard](dashboard.qmd) red-light card.

Only [Factor1 and Factor4]{.verdict-green} are wired as survival predictors, and
they were chosen **outcome-blind** — on platform-cleanliness and variance
explained alone. Choosing factors on survival association would bias the model
and is forbidden.

```{r}
#| label: platform-table
#| echo: false
if (!has_store) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the per-factor variance and platform-AUC table")
  )
} else {
  labs <- fn_factor_platform_labels(rownames(mofa_varexp), factor_platform)
  ve <- round(as.data.frame(mofa_varexp[labs$factor, , drop = FALSE]), 2)
  names(ve) <- paste0(names(ve), " var %")
  knitr::kable(
    data.frame(
      Factor         = labs$factor,
      ve,
      `Platform AUC` = round(labs$auc, 3),
      `BH q`         = signif(labs$q_value, 3),
      # fn_verdict_span() + escape = FALSE: the prose above shows CONFOUNDED as
      # [CONFOUNDED by assay]{.verdict-red}, and an unstyled verdict column
      # under it would say the opposite of what that sentence promises.
      Verdict        = fn_verdict_span(labs$verdict),
      `Survival predictor` =
        ifelse(labs$factor %in% PLATFORM_CLEAN_MOFA_FACTORS, "yes", "no"),
      check.names = FALSE, stringsAsFactors = FALSE
    ),
    escape = FALSE, row.names = FALSE
  )
}
```

**One caveat on the `BH q` column, because two public places print it
differently.** The table above renders the q values exactly as the DAG computes
them. `Factor6`'s Wilcoxon *p* underflowed to exactly 0 — impossible at
n = 214/310, where the normal approximation floors around 1.98e-84 — and because
BH ranks on *p*, that bad value shifts every larger *p* one rank down and
depresses the q of the factors below it by at most 2×. So `Factor2` prints
**1.4e-50** here against a corrected **2.9e-50**, and `Factor5` prints
**1.3e-34** against **2.0e-34**; the design spec and `R/constants.R` quote the
corrected pair. Nothing downstream changes: the platform-clean set at
`q > 0.05` is identical either way, which is why both numbers are allowed to
stand rather than one being quietly dropped. The arithmetic is worked out in
the post-hoc annotation of
`docs/results/platform-diagnosis-run-30911448546.txt`.

## Variance explained per omics view, per factor

Each factor is a shared axis of variation; the heatmap shows how much variance
it explains in each view. The row labels carry the platform AUC, so a
methylation-heavy row cannot be read without its confound.

```{r}
#| label: varexp
if (!has_store) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the variance-explained heatmap")
  )
} else {
  labs <- fn_factor_platform_labels(rownames(mofa_varexp), factor_platform)
  ve_long <- data.frame(
    factor   = rep(labs$label, times = ncol(mofa_varexp)),
    view     = rep(colnames(mofa_varexp), each = nrow(mofa_varexp)),
    variance = as.numeric(mofa_varexp),
    stringsAsFactors = FALSE
  )
  # Factor1 at the top, Factor15 at the bottom: the label order, not
  # alphabetical order, which would put Factor10 second.
  ve_long$factor <- factor(ve_long$factor, levels = rev(labs$label))
  p_var <- ggplot(ve_long, aes(x = view, y = factor, fill = variance)) +
    geom_tile(colour = "white") +
    scale_fill_viridis_c(name = "Var. (%)") +
    labs(x = "Omics view", y = "MOFA factor  [platform AUC, verdict]") +
    theme_minimal(base_size = 11)
  ggplotly(p_var)
}
```

## Sample scores: a clean axis against a confounded one

`Factor1` (platform-clean) is plotted against `Factor2` (the most
platform-loaded factor), **coloured by methylation assay**. If the two colours
separate along the x-axis, that separation is the assay, not the disease. This
plot exists to make the confound visible rather than to hide it — which is why
`Factor2` appears here and nowhere in the survival model.

```{r}
#| label: scores
if (is.null(mofa_factors) || is.null(methyl_platform)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the factor-score scatter")
  )
} else {
  ids <- rownames(mofa_factors)
  df <- data.frame(
    Factor1  = mofa_factors[, "Factor1"],
    Factor2  = mofa_factors[, "Factor2"],
    platform = as.character(methyl_platform[ids]),
    stringsAsFactors = FALSE
  )
  plot_ly(df, x = ~Factor1, y = ~Factor2, color = ~platform,
          colors = DASHBOARD_PLATFORM_COLOURS,
          type = "scatter", mode = "markers",
          marker = list(size = 6, opacity = 0.7)) |>
    layout(xaxis = list(title = "Factor 1 (platform-clean)"),
           yaxis = list(title = "Factor 2 (CONFOUNDED by assay)"),
           legend = list(title = list(text = "Methylation assay")))
}
```

## What the factors track — and what that is worth

```{r}
#| label: mutation-annot
#| echo: false
mfa <- fn_dashboard_read("mutation_factor_annot")
if (is.null(mfa)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the factor-vs-driver-mutation association table")
  )
} else {
  top <- head(mfa[order(mfa$q_value), ], 10)
  knitr::kable(top, row.names = FALSE, digits = 4)
}
```

Driver-mutation status was tested against every factor as an **external label**
on the **n = 413** cases where the 417-case mutation subset meets the 524-case
main cohort (`fn_annotate_mutation()` intersects the two before testing). Exactly **one** association survives BH correction:
**BAP1 ↔ Factor14, q = 0.004**. Both halves of that sentence matter:

- it is a real, corrected association, and it is the only one; **and**
- **Factor14 is a minor axis** — it explains 0.22 % of RNA, 0.31 % of
  methylation and 1.91 % of CNV variance. A significant association with a
  factor that explains almost nothing is a weak result, not a headline. Factor14
  also carries some platform signal itself (AUC 0.578, q 0.0046).

So this is reported as "the mutation signal that exists is confined to a minor
axis", never as "MOFA recovered BAP1 biology".

::: {.dashboard-note}
Factors are the frozen MOFA2 output; mutation status (BAP1/PBRM1) is used only
to *interpret* which factor it tracks — mutation is never a MOFA view. Recorded
values for the platform diagnostic are committed at
`docs/results/platform-diagnosis-run-30911448546.txt` (run 30911448546) and for
the factor/variance/mutation tables at
`docs/results/module2-run-30718392588.txt` (run 30718392588); this page renders
the same quantities live from the store rather than restating them.
:::
