---
title: "Reproducible Multi-Omics for Kidney Cancer (TCGA-KIRC)"
---

```{r}
#| label: setup
#| include: false
source("../R/constants.R")
source("../R/functions_dashboard.R")

# Same guard as every other page: NULL when the `targets-store` release asset
# has not been restored, so the tables below degrade to a stated gap instead of
# taking the landing page down. The plan calls `targets::tar_read_raw()`
# directly here, which errors on a fresh clone.
subtypes_df      <- fn_dashboard_read("subtypes_df")
survival_metrics <- fn_dashboard_read("survival_metrics")
bap1_auroc       <- fn_dashboard_read("bap1_auroc")
sanity_table     <- fn_dashboard_read("sanity_table")
```

::: {.provenance .provenance-frozen}
[Frozen]{.provenance-label} Every number on this page comes from the **frozen
`curatedTCGAData` KIRC snapshot 20160128 (hg19), data version 2.0.1**. It never
updates: re-running the pipeline reproduces it byte-for-byte. The project's only
live numbers are on the [Live GDC](live-gdc.qmd) page, and a snapshot count is
**not** interchangeable with a current GDC portal count — the two differ
materially in both directions.
:::

A portfolio project demonstrating **reproducible somatic multi-omics
integration** on public kidney-cancer data: `curatedTCGAData` → MOFA2 / SNF
integration → literature positive controls → a low-dimensional survival model
and a non-circular BAP1-from-expression classifier → this Quarto/Plotly
dashboard, orchestrated by `targets` and pinned by `renv` + Docker.

It is also a record of two negative results that were kept rather than tuned
away. Read the [Dashboard](dashboard.qmd) red-light card and the
[Survival](survival.qmd) qualifications before quoting anything here.

## Skills → module map

Each row names the code that implements the skill **and the run that executed
it**. A row with no run behind it says so; nothing in this table is claimed on
the strength of an intention. Four runs have their raw output committed under
`docs/results/` and can be opened: `30718392588`, `30840373033`, `30911448546`,
`31375702141`. The other two ids below are marked *(transcript not committed)*,
because a run id a reader cannot open is weaker evidence than one they can.

| Skill | Where it lives | Status — and the evidence |
|---|---|---|
| Somatic multi-omics integration | `R/functions_integrate.R` (MOFA2 + SNF), Modules 1–3 | **Run on real data.** 15 factors on n = 524, four subtypes (20 / 306 / 76 / 122), MOFA-vs-SNF ARI 0.351 — run `30718392588`, re-emitted by `31375702141` |
| Transcriptomic / epigenomic / CNV curation | `R/functions_ingest.R`, `R/functions_preprocess.R` | **Run on real data.** `rna_mat` 5000 × 524, `methyl_mat` 5000 × 524 (HM27 ∪ HM450 on common CpGs), `cnv_mat` 24776 × 524, `mut_annot` 417 × 7 — run `30708943504` (transcript not committed) |
| Reproducible pipeline + version control + environment manager | `_targets.R`, `renv.lock` (218 packages), `Dockerfile`, seven GitHub Actions workflows | **Run.** Every result above was produced inside `bioconductor/bioconductor_docker:RELEASE_3_23`, with the research packages installed by `BiocManager`. `renv.lock` pins the environment the `Dockerfile` restores and `ci.yml` asserts against, not the one these figures came out of |
| R + Python in one pipeline | MOFA2 via `reticulate`/`basilisk` (Module 2); `python/bap1_classifier.py` via `reticulate` (Module 4) | **Run on real data.** MOFA2 trains against the container's system Python with no conda download (run `30570220145`, transcript not committed); the classifier's AUROC is below |
| ML + survival analysis | `R/functions_survival.R` (Cox / penalised Cox / RSF), `R/functions_model_eval.R` (C-index + calibration written from scratch) | **Fitted on real data** — run `31375702141`. Read the "what these do not mean" note below before quoting the C-index |
| Literature positive controls as tests | `R/functions_sanity.R`, `tests/testthat/test-sanity.R` | **Run on real data.** 24 anchor tests (201 expectations green, 2 red); 4 of 5 checks pass, **the 4-cluster methylation check fails and is pinned as an expected red** — run `31375702141` |
| Data viz (Plotly) + communicating to a non-technical reader | `dashboard/*.qmd`, this site | **Built and rendering.** Six pages; every card degrades to a stated gap when a target is missing |
| Public data → regularly updated tool | `R/functions_gdc_live.R`, `.github/workflows/cron.yml` | **BUILT, NOT YET RUN.** The weekly cron is wired to refresh the [live GDC panel](live-gdc.qmd) only, but it has not yet executed in its current form and needs a `targets-store` release asset that does not exist yet. Deployment of the rendered site is a separate manual action |
| Single-cell | `python/singlecell_qc.py`, `python/singlecell_annotate.py`, `R/functions_purity.R`, flag-gated Module 6 targets | **BUILT, FLAG-GATED OFF, NEVER RUN.** Module 6 (GSE159115 + the ESTIMATE purity gate) is v1.1 and non-blocking. The code exists and is unit-tested on synthetic fixtures only: `run_singlecell` is `false` everywhere, the GSE159115 pull has never been performed, no purity verdict has been computed, and no single-cell result is claimed anywhere — see the [Single-cell (v1.1)](singlecell.qmd) page's own gaps |
| Publication-quality static figures | — | **NOT BUILT.** The figures here are interactive HTML; there is no manuscript-figure export path in the repo |

## Headline results

```{r}
#| label: summary
#| echo: false
knitr::kable(data.frame(
  Metric = c("Multi-omics cohort n (frozen snapshot)",
             "MOFA subtypes",
             "Survival C-index, held-out (Cox)",
             "BAP1-from-expression AUROC, held-out",
             "Literature checks passed"),
  Value  = c(
    fn_metric_text(if (is.null(subtypes_df)) NULL else nrow(subtypes_df), "%d"),
    fn_metric_text(if (is.null(subtypes_df)) {
      NULL
    } else {
      length(unique(subtypes_df$subtype))
    }, "%d"),
    fn_metric_text(survival_metrics$cindex$cox),
    fn_metric_text(bap1_auroc$heldout_auroc),
    if (is.null(sanity_table)) {
      DASHBOARD_PENDING_VALUE
    } else {
      sprintf("%d / %d", sum(sanity_table$passed), nrow(sanity_table))
    }
  ),
  check.names = FALSE, stringsAsFactors = FALSE
), row.names = FALSE)
```

```{r}
#| label: summary-gap
#| echo: false
if (is.null(survival_metrics)) {
  htmltools::div(class = "pending-note",
                 fn_dashboard_pending("the headline metrics above"))
}
```

The values the restored store carries, recorded by GitHub Actions run
`31375702141` (2026-08-10) and transcribed in
[`docs/results/module4-run-31375702141.txt`](https://github.com/Lexiyao/kidney-cancer-multiomics/blob/main/docs/results/module4-run-31375702141.txt):
held-out C-index **Cox 0.7486 / penalised Cox 0.7492 / RSF 0.7524**, Cox
optimism (apparent − held-out) **0.0125**; BAP1-from-expression AUROC **0.960
held-out, 0.958 cross-validated** on n = 413 with 36 mutants.

### What those two numbers do **not** mean {.unnumbered}

**The C-index is a stage-and-age model.** In the fitted Cox model only the
clinical terms reach significance — `stage_num` *p* = 9.6 × 10⁻¹⁷,
`age_years` *p* = 9.7 × 10⁻⁶ — while **both wired MOFA factors do not**:
`Factor1` HR 0.982 (*p* = 0.60) and `Factor4` HR 0.966 (*p* = 0.24). A C-index
of 0.75 is therefore **not** evidence that multi-omics factors predict survival
in this cohort. On this data they add nothing detectable to stage and age, and
that is the finding.

**It is also not fully out-of-sample.** MOFA was fitted on all 524 cases and the
5000-gene variance filter was computed over all samples before either model drew
its train/test split. Both steps are outcome-blind — no label leakage — but the
held-out rows helped define the latent axes and the feature set, so the figures
are *unsupervised-transductive*. The 0.0125 optimism bounds the supervised
component only. There is no external validation cohort.

**The BAP1 AUROC rests on 36 mutants.** 36 of 413 cases carry a BAP1 mutation,
so the held-out split contains only a handful of positives and the interval
around 0.96 is wide. The task is genuinely non-circular — the label is an
external mutation call, not something derived from the expression it is
predicted from — but the point estimate should be read as "a strong expression
signature is plausibly there", not as a validated classifier.

## Literature positive controls (the credibility anchor)

```{r}
#| label: sanity
#| echo: false
if (is.null(sanity_table)) {
  htmltools::div(class = "pending-note",
                 fn_dashboard_pending("the positive-control results table"))
} else {
  knitr::kable(
    data.frame(
      Check    = sanity_table$check,
      Verdict  = fn_pass_span(sanity_table$passed),
      Evidence = sanity_table$detail,
      check.names = FALSE, stringsAsFactors = FALSE
    ),
    escape = FALSE, row.names = FALSE
  )
}
```

Four of the five checks pass on the real snapshot; **the merged-methylation
4-cluster check fails**. That partition tracks the **assay platform** (adjusted
Rand index 0.583 against a 0.25 veto), and four clusters are not recovered
inside either platform on its own — within-arm silhouettes HM27 0.0858 and
HM450 0.0489 are both *below* the merged 0.1197, so the merge manufactures the
structure rather than diluting it. The failure is pinned as an expected red in
the container run (`verify-module2.yml`): a *new* red fails the job, and so does
this anchor starting to pass. See [MOFA Factors](factors.qmd) for the per-factor
platform AUCs.

**What this red is not.** It is not a failure to reproduce published
methylation biology. TCGA KIRC (Nature 2013) identifies m1–m4 as four stable
mRNA **expression** subtypes and reports no four DNA-methylation strata for
KIRC; `k = 4` here is a design choice carried over from those expression
subtypes. The finding is about this merged HM27/HM450 matrix, not about the
paper.

What the confound does not touch: the MOFA **subtypes** are platform-clean (ARI
0.0058, *p* = 0.53), and the mutation-frequency (VHL 44.8 %, PBRM1 30.5 %,
SETD2 10.1 %, BAP1 8.6 %) and ccA/ccB (rho −0.354) anchors read no methylation
matrix at all.

::: {.dashboard-note}
**Honest scope.** Effective *n* is 241–524, set by modality intersection. The
research core is a frozen 2016 / hg19 snapshot and never updates. CI lints and
runs unit tests on subsampled fixtures; the deploying render is designed to read
a cached `_targets` store restored from a release asset, and that asset has not
been produced yet — **a green CI badge is not end-to-end reproduction**. Full limitations are in the
[README](https://github.com/Lexiyao/kidney-cancer-multiomics#known-limitations-stated-openly)
and `docs/runtime.md`.
:::
