---
title: "Live GDC Panel"
format:
  html:
    page-layout: full
---

```{r}
#| label: setup
#| include: false
library(plotly)
source("../R/constants.R")
source("../R/functions_dashboard.R")
source("../R/functions_gdc_live.R")

# `gdc_live_panel` is the ONE target in the DAG that is not frozen: it carries
# tar_cue(mode = "always") and re-queries api.gdc.cancer.gov on every run.
# NULL here means the store was not restored for this render, exactly as on
# every other page.
panel <- fn_dashboard_read("gdc_live_panel")
```

::: {.provenance .provenance-live}
[Live]{.provenance-label} **This is the only page in the project whose numbers
change.** Everything here is queried from the **GDC REST API**
(`api.gdc.cancer.gov/cases`) and describes TCGA-KIRC **as the GDC portal holds
it today**. Nothing on this page comes from the frozen research snapshot, and
**no number here is comparable to a cohort figure** on the
[Dashboard](dashboard.qmd), [MOFA Factors](factors.qmd) or
[Survival](survival.qmd) pages — those are all frozen 20160128 (hg19).
:::

## What updates, and what does not

The multi-omics research core is a **frozen 2016 `curatedTCGAData` snapshot**.
Re-running it weekly would reproduce it byte-for-byte, so claiming the *project*
is a regularly updated tool would be false. What is genuinely scheduled is this
panel: the `gdc_live_panel` target re-queries the GDC API on every pipeline run,
and the weekly cron (`.github/workflows/cron.yml`) is **wired** to re-run that
target and re-freeze the cached store. "Wired", not "does" — that cron has not
yet executed in its current form, and it restores a `targets-store` release
asset that does not exist yet, so it is a built mechanism rather than observed
behaviour.

**Honest scope of the "weekly" claim.** The cron refreshes the panel and the
cached store; publishing the rendered site is currently a **manual** action
(`.github/workflows/pages.yml` is `workflow_dispatch` only). So the number below
is current as of the **retrieval** timestamp it carries — the moment the
pipeline last queried the GDC, which can be weeks before this page was rendered,
and is certainly not the moment you are reading it. A deploy restores a cached
store and rebuilds nothing, so the render can be far newer than the retrieval.
The timestamp is printed with every figure for exactly that reason.

```{r}
#| label: total
#| echo: false
if (is.null(panel)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the live GDC case total and clinical facets")
  )
} else {
  htmltools::div(
    class = "provenance provenance-live",
    htmltools::HTML(sprintf(
      paste0("<strong>%d</strong> TCGA-KIRC cases at the GDC ",
             "(project <code>%s</code>), retrieved <strong>%s UTC</strong>."),
      panel$total_cases, panel$project_id,
      format(panel$retrieved_at, "%Y-%m-%d %H:%M", tz = "UTC")
    ))
  )
}
```

::: {.dashboard-note}
**The live GDC case total is not the analysis cohort.** It is every TCGA-KIRC
case the GDC currently indexes, on any assay or none. The analysis cohort is
`r COHORT_SIZES$rna_methyl_cnv` cases — those with RNA **and** methylation
(HM27 ∪ HM450) **and** CNV in the frozen 20160128 snapshot — falling to
`r COHORT_SIZES$rna_methyl_cnv_mutation` once mutation annotation is required.
The two sources disagree in both directions: the 2016 legacy MAF covers
`r COHORT_SIZES$mutation_subset` cases while today's masked-somatic MAF covers
fewer, and today's portal indexes cases the 2016 snapshot did not. Quoting a
live figure as a snapshot figure (or the reverse) is a silent error this project
has already made once and corrected.
:::

## Current clinical distribution at the GDC

Case counts by clinical field, **live**, for the whole TCGA-KIRC project. These
are portal-wide distributions, not the distribution of the analysis cohort, and
they are shown as counts rather than percentages so nothing implies a modelled
quantity.

```{r}
#| label: facets
#| echo: false
if (is.null(panel)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the live clinical-distribution charts")
  )
} else {
  # Built as ONE tagList rather than printed inside a `results: asis` loop:
  # htmlwidgets emitted from a loop lose their dependency registration and the
  # charts come out blank. Assembling the headings and widgets together lets
  # knitr collect the plotly dependencies once, for all of them.
  htmltools::tagList(lapply(names(panel$facets), function(field) {
    df <- panel$facets[[field]]
    fig <- plot_ly(df, x = ~n, y = ~reorder(category, n), type = "bar",
                   orientation = "h",
                   marker = list(color = DASHBOARD_ACCENT)) |>
      layout(xaxis = list(title = "Cases (live GDC)"),
             yaxis = list(title = ""),
             margin = list(l = 160))
    # The stamp goes on EVERY chart, not just the total above: a facet chart
    # screenshotted out of this page must still say when the portal was queried,
    # because the store it renders from can be weeks older than the render.
    htmltools::tagList(
      htmltools::h3(fn_gdc_facet_title(field)),
      htmltools::div(class = "provenance provenance-live",
                     fn_gdc_facet_note(field, panel$retrieved_at)),
      fig
    )
  }))
}
```

::: {.provenance .provenance-live}
[Live]{.provenance-label} A facet that GDC stops serving is treated as drift,
not as an empty result: `fn_extract_gdc_buckets()` stops with the API's own
warning rather than rendering a blank chart, so the weekly cron fails loudly
instead of publishing a panel that quietly lost a field. That is not
hypothetical — the field this page shows as *Sex at birth* was
`demographic.gender` when it was first written, and GDC no longer defines it.
:::
