---
title: "Single-cell increment (v1.1) — GSE159115"
format: html
---

```{r}
#| label: setup
#| include: false
source("../R/constants.R")
source("../R/functions_dashboard.R")

# Read through the OPTIONAL guard, not fn_dashboard_read(): every target on
# this page is either declared behind the `run_singlecell` flag (Task 6.8) or
# was added by Module 6 after the release store was frozen, so "the store is
# present and has no such target" is the expected state here and must degrade
# to a stated gap rather than abort the site render.
#
# The plan's version of this chunk calls `tar_manifest()` and a bare
# `tar_read()`. Neither works from a rendering page: `_quarto.yml` sets
# `execute-dir: file`, so the working directory is dashboard/, where there is
# no `_targets.R` for tar_manifest to parse and no `_targets` store for
# tar_read to find. Both would take the whole render down.
purity_check <- fn_dashboard_read_optional("purity_check")
sc_mapping   <- fn_dashboard_read_optional("sc_mapping")
```

::: {.callout-warning}
## Non-blocking increment
This page is v1.1. It is built on a separate branch and does **not** gate
release of the multi-omics core (Modules 0–5). Nothing on it is a finding: the
GSE159115 single-cell data has not been processed anywhere yet, and no
scientific claim is made from it on this site.
:::

GSE159115 is the single-cell RNA-seq series of Zhang Y, Narayanan SP,
Mannan R, *et al.*, "Single-cell analyses of renal cell cancers reveal
insights into tumor microenvironment, cell of origin, and therapy response,"
*Proc Natl Acad Sci U S A* 2021;118(24) (PMID
[34099557](https://pubmed.ncbi.nlm.nih.gov/34099557/), DOI
[10.1073/pnas.2103240118](https://doi.org/10.1073/pnas.2103240118)) — cited
here the same way ESTIMATE carries Yoshihara *et al.* in `R/constants.R`,
because a dataset used without attribution is a claim without a source.

## Why there is a gate before there is a mapping

Mapping bulk subtypes onto single cells is only interesting if the subtypes are
something other than a **tumour-purity / immune-infiltration** readout. In
ccRCC that is a live possibility, not a formality: if the subtypes largely
track how much tumour (versus stroma and immune infiltrate) is in each bulk
sample, then scoring single cells against their signatures rediscovers
*cell-type proportion* and nothing about a cellular basis for the subtypes.

The gate has **two arms**, one per confound: Kruskal-Wallis of `TumorPurity ~
subtype` and of `ImmuneScore ~ subtype`, each with an epsilon-corrected
eta-squared. It fires when **either** arm is both significant (p < 0.05) and
large (eta-squared ≥ 0.14), and the verdict records which arm fired. An OR and
not an AND: immune infiltration is the more likely of the two confounds in
ccRCC, so a gate reading only the purity arm would clear a cohort whose
subtypes are a pure immune proxy.

So the check runs **first**, on bulk data alone. `purity_check` has no
single-cell input and no edge to any single-cell target, and the mapping takes
its verdict as an argument — `map_bulk_signature(..., purity_confounded =
purity_check$is_purity_proxy)` — which refuses a missing or non-boolean verdict
rather than reading it as a pass. The gate can return either answer: both arms
are asserted in `tests/testthat/test-purity-targets.R` against independent
synthetic data, because a gate that cannot fail licenses the mapping
unconditionally and is the same defect as having no gate at all.

## Purity confound gate

```{r}
#| label: purity-gate
#| echo: false
if (is.null(purity_check)) {
  htmltools::div(
    class = "pending-note",
    htmltools::HTML(fn_singlecell_gap("the purity confound-gate verdict"))
  )
} else {
  htmltools::div(
    class = "provenance provenance-frozen",
    htmltools::HTML(fn_purity_gate_line(purity_check))
  )
}
```

::: {.dashboard-note}
**The purity figure is an approximation, and it is only used as a confound
probe.** Tumour purity here is the ESTIMATE transformation of Yoshihara et al.
(*Nat Commun* 2013), whose two coefficients were calibrated on Affymetrix
arrays against ABSOLUTE. TCGA-KIRC RNA-seq is not that platform, so the value
is never reported as a measured purity — it is a covariate used to ask one
question about the subtypes and nothing else.
:::

## Bulk → single-cell mapping

```{r}
#| label: mapping-interpretation
#| echo: false
if (is.null(sc_mapping)) {
  htmltools::div(
    class = "pending-note",
    htmltools::HTML(fn_singlecell_gap("the bulk to single-cell mapping"))
  )
} else {
  htmltools::div(
    class = "provenance provenance-frozen",
    htmltools::HTML(sc_mapping$interpretation)
  )
}
```

```{r}
#| label: mapping-table
#| echo: false
# A SEPARATE chunk from the interpretation above, not decoration: `kable()`
# returns a multi-element character vector, and htmltools::tagList() cannot
# hold one ("Text to be written must be a length-one character vector"), so
# emitting both from one chunk aborts the whole site render.
if (!is.null(sc_mapping)) {
  knitr::kable(
    fn_sc_score_table(sc_mapping),
    digits = 3,
    caption = paste(
      "Mean single-cell score per cell type for each bulk subtype",
      "signature. Signatures are the top", SC_SIGNATURE_N_GENES,
      "genes by subtype-versus-rest mean difference -- a ranking, NOT",
      "differential expression: no test, no multiple-testing correction,",
      "no fold-change floor. Each column header carries the signature's gene",
      "coverage (genes found in the single-cell data / genes requested),",
      "because the signatures are derived on bulk TCGA expression and partial",
      "dropout in the scRNA gene space is the expected case. NaN marks a",
      "signature with no measurable gene; it is never rendered as 0."
    )
  )
}
```

## What the labels on those cells are, and are not

Cell types are assigned by `argmax` over per-cell marker scores across six
canonical sets — ccRCC tumour epithelium (`CA9`, `NDUFA4L2`, `VEGFA`, `EPCAM`),
T cells (`CD3D`, `CD3E`, `CD8A`, `IL7R`), myeloid (`CD14`, `LYZ`, `CD68`),
endothelium (`PECAM1`, `VWF`, `CLDN5`), fibroblast/pericyte (`COL1A1`, `ACTA2`)
and B cells (`CD79A`, `MS4A1`). Before the `argmax`, each score is
standardised against a **permutation null** of same-size random gene sets:
raw `score_genes` outputs are not comparable across marker sets of different
size — the set-mean's sampling variance scales as 1/(panel size), so on pure
noise the 2-gene panels won a raw argmax 1.76× as often as the 4-gene ones.
The standardisation removes that size bias; it does **not** add a confidence
threshold.

Five limitations travel with every label:

- There is **no confidence threshold**. Every cell receives one of the six
  labels whether or not it belongs to any of them.
- There is **no proximal-tubule set**, and `EPCAM` is frequently low in ccRCC,
  so adjacent-normal tubule cells can be pulled into `Tumor_epithelial`.
- `ACTA2` also marks pericytes and vascular smooth muscle, so the fibroblast
  set is not pure.
- `NDUFA4L2`, in the tumour set, is also a documented **pericyte** marker
  (Mesa-Ciller et al., *J Cereb Blood Flow Metab* 2023, PMID 35929074).
  ccRCC is among the most vascularised solid tumours, so pericyte signal is
  split between two labels — the `ACTA2` fibroblast set and the tumour set
  itself.
- `VEGFA` is a hypoxia-response gene expressed broadly by myeloid, stromal
  and tubular cells, not a tumour-cell-restricted marker, so angiogenic
  macrophages are pulled toward `Tumor_epithelial`.

## Reproduction

Nothing on this page runs in CI or in the frozen release. `run_singlecell` is
`false` in `config/params.yml`, so the single-cell targets are not even
declared on a runner and the GSE159115 download cannot start there. Running it
locally means flipping that flag in a local override
(`R_CONFIG_FILE=config/params.local.yml`), supplying the data, and building
`sc_mapping` in the container, where `reticulate`, `scanpy` and the R-Forge
`estimate` package exist.

"Supplying the data" is not one file from GEO: the series' supplementary
artefact is `GSE159115_RAW.tar`, a tar of **per-sample** 10x H5 files covering
14 specimens — 8 renal tumours and 6 benign kidneys, spanning ccRCC **and**
chromophobe RCC — plus per-histology annotation CSVs. The wiring as written
processes exactly **one** specimen's H5 (there is no concatenation, no
per-sample column, no batch handling), so `h5_path` in `config/params.yml`
must be pointed at one extracted per-sample file, and any result would be
about that single specimen, not the series. The committed `h5_path` is null
and the `sc_h5_path` target refuses to run until it is set. The scanpy
helpers are unit-tested against a synthetic 10x fixture; the GSE159115 pull
itself has not been performed.
