---
title: "Survival"
format:
  html:
    page-layout: full
---

```{r}
#| label: setup
#| include: false
library(plotly)
library(survival)
source("../R/constants.R")
source("../R/functions_dashboard.R")

# Frozen research targets; NULL when the release-asset store has not been
# restored, in which case each section prints a stated gap instead of a number.
km_src           <- fn_dashboard_read("km_subtype_df")
survival_metrics <- fn_dashboard_read("survival_metrics")
cox_fit          <- fn_dashboard_read("cox_fit")
```

::: {.provenance .provenance-frozen}
[Frozen]{.provenance-label} Every number on this page comes from the **frozen
`curatedTCGAData` snapshot 20160128 (hg19)**. Outcomes, follow-up and the fitted
model **never update**; a weekly cron re-running this would reproduce it
byte-for-byte. The only live-updating numbers in the project are on the
[Live GDC](live-gdc.qmd) page.
:::

## What is being fitted

A deliberately **low-dimensional** overall-survival model:
`Factor1`, `Factor4`, `age_years`, `stage_num`, `platform`. Five terms - well
inside the EPV-10 budget the observed event count licenses, and never
genome-wide feature selection.

`Factor1` and `Factor4` were chosen **outcome-blind**, on platform-cleanliness
(BH `q > 0.05` against the HM27/HM450 split) and variance explained only. The
two largest methylation-loaded factors, `Factor2` (AUC 0.888) and `Factor5`
(0.818), are substantially the assay split and are **excluded**; `platform`
enters instead as an explicit adjustment covariate. See
[MOFA Factors](factors.qmd). Selecting factors on survival association would
bias the model and is forbidden.

The same assay merge also fails a positive control: the merged-methylation
4-cluster anchor is [RED]{.verdict-red} (cluster-vs-platform ARI 0.583 against a
0.25 veto) and stays red. See the [Dashboard](dashboard.qmd) red-light
card.

```{r}
#| label: model-size
#| echo: false
if (is.null(cox_fit)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the train/test split, event counts and EPV budget")
  )
} else {
  knitr::kable(fn_cox_fit_summary(cox_fit), col.names = c("", ""),
               row.names = FALSE)
}
```

## Kaplan–Meier overall survival by MOFA subtype

**Descriptive only.** The Cox model above has **no `subtype` term**, so these
curves are not the model's output and no *p*-value from them is a model result.
They are built from the `km_subtype_df` join, not from the fitted object. The
MOFA subtypes are, at least, platform-clean (ARI 0.0058 against the assay
split), so this particular figure is not a picture of the confound.

```{r}
#| label: km
if (is.null(km_src)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the Kaplan-Meier curves by MOFA subtype")
  )
} else {
  km <- survfit(Surv(time, status) ~ subtype, data = km_src)
  # fn_km_curve_df() is shared with dashboard.qmd so the two pages cannot draw
  # the same curves from two slightly different expansions of the same fit.
  plot_ly(fn_km_curve_df(km), x = ~time, y = ~surv, color = ~strata,
          type = "scatter", mode = "lines",
          line = list(shape = "hv")) |>
    layout(xaxis = list(title = "Time (days)"),
           yaxis = list(title = "Overall survival", range = c(0, 1)))
}
```

## Model evaluation (reused model-evaluation-from-scratch)

The **held-out** C-index is the number to read. The **optimism** column beside
it - apparent (training) minus held-out - is *reported, not subtracted away*:
the factors and subtypes were derived from the same omics the model is fitted
on, so the apparent figure is optimistic by construction and the size of that
gap is itself the finding. Only the Cox arm has an optimism estimate; the other
rows are left blank rather than borrowing it.

```{r}
#| label: metrics
#| echo: false
if (is.null(survival_metrics)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending(paste(
      "the held-out C-index (Cox / penalised Cox / RSF) and the Cox optimism"
    ))
  )
} else {
  ct <- fn_survival_cindex_table(survival_metrics)
  knitr::kable(
    data.frame(
      Model                          = ct$arm,
      `Held-out C-index`             = round(ct$heldout, 3),
      `Optimism (apparent - held-out)` = round(ct$optimism, 3),
      `Apparent (training) C-index`  = round(ct$apparent, 3),
      check.names = FALSE, stringsAsFactors = FALSE
    ),
    row.names = FALSE
  )
}
```

::: {.provenance .provenance-frozen}
[Caveat]{.provenance-label} **The held-out C-index is not fully out-of-sample
either, and that is stated here rather than in a footnote.** `mofa_factors` was
fitted by MOFA2 on **all 524 cases** - the same rows the train/test split later
calls TEST - so the latent axes the model is built on were defined with the
held-out rows contributing. MOFA never sees an outcome, so this is **not a label
leak** and the non-circularity claim stands; it is **unsupervised-transductive
optimism**. The figure above therefore bounds the **supervised component only**.
A fully out-of-sample estimate would require refitting MOFA (and the
top-variable-gene filter) inside every training split, which is out of scope at
this cohort size. There is also **no external validation cohort**.
:::

## Calibration at the 5-year horizon

Predicted 5-year survival against observed, from `basehaz(centered = TRUE)` and
the linear predictor on one shared centring. Points on the dotted line are
well-calibrated; systematic departure from it is a miscalibrated model, and it
is shown whichever way it falls.

```{r}
#| label: calibration
if (is.null(survival_metrics) || is.null(survival_metrics$calibration)) {
  htmltools::div(
    class = "pending-note",
    fn_dashboard_pending("the 5-year calibration curve")
  )
} else {
  plot_ly(survival_metrics$calibration, x = ~predicted, y = ~observed,
          type = "scatter", mode = "markers+lines",
          marker = list(color = DASHBOARD_ACCENT)) |>
    add_lines(x = c(0, 1), y = c(0, 1),
              line = list(dash = "dot", color = "grey"),
              showlegend = FALSE) |>
    layout(xaxis = list(title = "Predicted 5-year survival", range = c(0, 1)),
           yaxis = list(title = "Observed",                  range = c(0, 1)))
}
```

::: {.dashboard-note}
**Limitations, stated on the page that would otherwise oversell.** Effective *n*
is set by modality intersection - 524 for the main RNA + Methylation + CNV
cohort, 413 once mutation annotation is required, 241 if HM450 is demanded
instead of the merged platforms; survival runs on the main cohort, with **519 of
the 524 cases reaching the fit** (171 OS events, 32.9 %) - the same 519 the
train 364 / held-out 155 split in the table above sums to.
**Events, not *n*, bound the model**, which is why it stays at five terms under
the EPV-10 rule. The BAP1 survival positive control is *directionally*
consistent with the literature (HR 1.584, 95% CI 0.967–2.595, *p* = 0.068) but
**underpowered by construction** - roughly 470 events would be needed for 80%
power, and the mutation subset records **147, 18 of them in the mutant arm**
(run 31375702141), about 3.2× short - so it is reported as neither a
confirmation nor a negative finding. Full limitations in the README.
:::
